Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-05 paper/paper.tex
added separate Competing Interests section per instructions
revisions to Intro and first section of Results (substantially shortened)

Unfortunately for the host, influenza infection only rarely triggers IFN production by infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

@@

Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four type of defects that cause increased IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich for rare IFN+ positive cells}
A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
This paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.

This rareness of IFN induction is not merely a feature of A549 cells in culture.
Prior studies using reporters have found that IFN expression is rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes (ISGs), which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
But just a handful of cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.28\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
\hspace{0.02\textwidth}
{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large C} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
IFN expression is rare in influenza-infected cells.
{\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
{\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\FIGDATA[IFNrare]{reporter_sequences}).
The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
{\bf (C)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}

\figsupp[Few cells express detectable IFN in influenza-infected mice.]
{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one transcript detected).
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
\label{figsupp:mice}

@@

To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and expanded cell lines from single clones.
These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
Cells that activate the IFN reporter can enriched by magnetic-activated cell sorting (MACS) for the cell-surface protein, or identified by flow cytometry for the fluorescent protein.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics described below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We used the reporter cells to quantify the frequency of IFN induction in A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
This frequency of IFN induction is roughly comparable to that observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B).

@@

\section{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, and AJ Velthuis for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
We thank Katherine Xue, David Bacsik, and Hannah Itell for comments on the manuscript.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.
The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
The authors declare that no competing interests exist.

>>>>

Unfortunately for the host, influenza infection only rarely triggers IFN production by infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

@@

Here we develop a new approach to measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four types of defects that cause increased IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, as even wild-type virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich for rare IFN+ positive cells}
A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}, and found that only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
Prior studies using reporters have found that IFN expression is also rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}, a conclusion supported by our own re-analysis of the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection} (see \FIGSUPP[IFNrare]{mice}).

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.28\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
\hspace{0.02\textwidth}
{\bf \Large B}\includegraphics[width=0.35\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large C} \includegraphics[width=0.23\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
IFN expression is rare in influenza-infected cells.
{\bf (A)} Number of infected and IFN+ A549 cells in our prior single-cell transcriptomics~\citep{russell2018extreme}.
{\bf (B)} An A549 reporter cell line to identify cells expressing IFN.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\FIGDATA[IFNrare]{reporter_sequences}).
The reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
{\bf (C)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}

\figsupp[Few cells express detectable IFN in influenza-infected mice.]
{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
\label{figsupp:mice}

@@

To efficiently identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein (\FIG{IFNrare}B).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in our cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of ``wild-type'' A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells---a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}A).

@@

\section{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, and AJ Velthuis for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
We thank Katherine Xue, David Bacsik, and Hannah Itell for comments on the manuscript.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.
The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.

\section{Competing Interests}
The authors declare that no competing interests exist.
